---
title: "PS270 Final Project"
author: "Carson Phillips"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
```{r}
library(tidyverse)
inflation <- read_csv("data/inflation.csv", show_col_types = FALSE)
presapp <- read_csv("data/presapp270.csv", show_col_types = FALSE)

head(inflation)
head(presapp)
```



## Project thoughts

I am interested in exploring data related to economics and political science. For example, I would be curious to research whether higher inflation is correlated with lower presidential approval ratings.

## Research Hypothesis

My research question is as follows: does higher inflation have an impact on presidential approval ratings? 
My hypothesis that higher inflation is associated with lower presidential approval ratings. I believe that this is the case because the public generally pays attention to inflation as a general indicator of economic health, and also attributes economic outcomes to the sitting president. By extension, they should disapprove more of a president who is in office while inflation is higher. My explanatory variable of interest is yearly inflation from 1960 - 2024, sourced from the Federal Reserve Bank of St. Louis "Inflation, consumer prices for the United States" indicator. This variable is measured annually and expressed as a percentage. My outcome variable of interest is yearly presidential approval ratings from time in office from Gallup, ranging from Harry Truman to Joe Biden and also expressed as a percentage. I recorded yearly presidential approval ratings from Gallup's "Presidential Job Approval Center" into an Excel document. If a given increase in inflation corresponds to a lower presidential approval rating, this would provide support for my hypothesis. A given increase in inflation has no positive or negative correlation on presidential approval ratings, or a given increase in inflation is associated with higher presidential ratings, would disprove my hypothesis.

## Data cleaning and visualization

```{r}
presapp <- presapp |>
  rename(year = Date) |>
  mutate(
    approval = parse_number(`Approval Rating`))

presapp

inflation <- inflation |>
  rename(inflation = FPCPITOTLZGUSA) |>
  mutate(year = as.integer(format(observation_date, "%Y")))

inflation
```



```{r}
library(tidyverse)

joined <- presapp |>
  inner_join(inflation, by = "year")

joined
```

```{r}
visual <- ggplot(data = joined, mapping = aes(x = inflation, y = approval, color = President)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1") +
  labs(x = "Yearly Inflation (%)",
       y = "Yearly Presidential Approval Rating (%)",
       title = "Relationship Between Inflation and Presidential Approval") +
  theme_minimal()

visual
```

## Regression
```{r}
library(broom)
approval_fit <- lm(approval ~ inflation, data = joined)

regression_table <- tidy(approval_fit)

regression_table |> 
  knitr::kable(
    digits = 2,
    col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"),
    caption = "Table 1: Linear Regression of Presidential Approval on Inflation"
  )
```
According to my regression analysis, the predicted presidential approval rating when yearly inflation is 0% is 57.43%. The slope is -1.27, meaning that every 1% increase in yearly inflation corresponds to a drop in presidential approval ratings of 1.27%. These findings support my hypothesis that higher inflation is associated with lower presidential approval ratings. One thing to note is the p-value of 3%, which informs us that there is an 3% chance these findings would be visible even if inflation had no effect on presidential approval ratings. 
