---
title: "PS270 Final Project"
author: "Carson Phillips"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
```{r}
library(tidyverse)
inflation <- read_csv("data/inflation.csv", show_col_types = FALSE)
presapp <- read_csv("data/presapp270.csv", show_col_types = FALSE)

head(inflation)
head(presapp)
```



## Introduction

**My research question is as follows**: does higher inflation have an impact on presidential approval ratings? **I hypothesize that higher inflation is associated with lower presidential approval ratings.** The **null hypothesis (H0)** in this case is that there is no relationship between yearly inflation and presidential approval ratings. The **alternative hypothesis (H1)** would be that higher inflation is associated with lower presidential approval ratings. I believe this relationship will be evident, because the public generally pays attention to inflation as a general indicator of economic health, and also attributes economic outcomes to the sitting president. By extension, they should disapprove more of a president who is in office while inflation is higher. 

**This research question is important**, because it seeks to provide explanation for presidential approval ratings, the primary indicator used to define the public's assessment of the president. Presidential approval ratings have a wide berth of consequences. They indicate the president's likelihood of reelection, the likelihood that president's party will find success at midterm elections, the level of difficulty the president will incur in trying to pass legislation, and other considerations. As a result, it's important to understand what variables impact presidential approval ratings. The rate of inflation is one of the most visible aspects of the economy in the public's eyes, and the public also looks to the president to generate economic success. Thus, it is logical to research whether the rate of inflation informs the public's opinion of the president. 

## Data

My explanatory variable of interest, or **predictor variable, is** yearly inflation from 1960 - 2024, sourced from the Federal Reserve Bank of St. Louis's "Inflation, consumer prices for the United States" indicator. This variable is measured annually and expressed as a percentage. **My outcome variable of interest is** Gallup's yearly presidential approval ratings from time in office, ranging from Harry Truman to Joe Biden, and also expressed as a percentage. I specifically recorded yearly presidential approval ratings from Gallup's "Presidential Job Approval Center" into an Excel document. If the results are statistically significant, and a given increase in inflation corresponds to a lower presidential approval rating, **this would provide support for my alternative hypothesis.** If a given increase in inflation has no correlation with presidential approval ratings, or if a given increase in inflation is associated with higher presidential ratings, **this would mean that I reject my hypothesis.** The **decision rule** would indicate that I should reject the **null hypothesis** if the regression coefficient for inflation is negative, and the p-value is less than 0.05 (since I will run this regression at the 95% Confidence Interval). According to the **rejection criteria**, I should will fail to reject the null hypothesis if the p-value is greater than 0.05, or if the regression coefficient for inflation is positive. 

## Data Cleaning

### Preview of Inflation and Presidential Approval Ratings Datasets

```{r}
presapp <- presapp |>
  rename(year = Date) |>
  mutate(
    approval = parse_number(`Approval Rating`))

presapp

inflation <- inflation |>
  rename(inflation = FPCPITOTLZGUSA) |>
  mutate(year = as.integer(format(observation_date, "%Y")))

inflation
```

### **Plot Visualizing Presidential Voting Records**

```{r}
presappplot <- ggplot(data = presapp, mapping = aes(x = year, y = approval, color = President)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1") +
  labs(x = "Year",
       y = "Presidential Approval Rating (%)",
       title = "Presidential Approval Ratings by Year")

presappplot
```

### Joining Datasets

```{r}
library(tidyverse)

joined <- presapp |>
  inner_join(inflation, by = "year") 

joined
```

## Results

### **Plot Visualizing Presidential Approval Ratings with Relation to Changes in Inflation**

```{r}
visual <- ggplot(data = joined, mapping = aes(x = inflation, y = approval, color = President)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1") +
  labs(x = "Yearly Inflation (%)",
       y = "Yearly Presidential Approval Rating (%)",
       title = "Relationship Between Inflation and Presidential Approval Ratings") +
  theme_minimal()

visual
```

## Regression
```{r}
library(broom)
approval_fit <- lm(approval ~ inflation, data = joined)

regression_table <- tidy(approval_fit)

regression_table |> 
  knitr::kable(
    digits = 2,
    col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value"),
    caption = "Table 1: Linear Regression of Presidential Approval on Inflation"
  )
```
According to my regression analysis, the predicted presidential approval rating when yearly inflation is 0% is 57.43%. The slope is -1.27, meaning that every 1% increase in yearly inflation corresponds to a drop in presidential approval ratings of 1.27%. These findings support my hypothesis that higher inflation is associated with lower presidential approval ratings. Furthermore, the p-value of 3% informs us that there is an 3% chance these findings would be visible even if inflation had no effect on presidential approval ratings. Referring back to the null and alternative hypotheses, these results propound that I can adopt the alternative hypothesis. The regression coefficient is negative, and the p-value is less than 0.05 (0.03). Since this analysis is run at the 95% confidence interval, a p-value lower than 0.05 would indicate statistically significant results. 